---
title: "Future Oak: AOD and COD comparison"
author: "Marine C. Cambon"
date: last-modified
published-title: "document last compilation"
format:
  html:
    theme: 
      light: minty
      dark: slate
    linestretch: 1.7
    toc: true
    toc-location: left
    code-fold: true
    code-summary: "Show the code"
    self-contained: true
    link-external-newwindow: true
    # comments:
    #   hypothesis: true
editor: source
#bibliography: documentation/references.bib
editor_options: 
  chunk_output_type: console
---

```{r}
rm(list = ls())
graph_par <- c(las=1, bty="l", col.axis="gray30")
hue_tissue <- c("Leaf" = "olivedrab3", "Stem" = "bisque2", "Soil" = "salmon4")
hue_disease <- c("Control" = "#abdda4", "DiseaseAOD" = "#d53e4f", "DiseaseCOD" = "#fdae61")

pdf_out <- TRUE
# Load custom functions from src
purrr::walk(list.files("src", pattern = ".R", full.names = T), source)
# renv::activate()

# Loading a few packages
library(phyloseq)
suppressMessages(library(dplyr))
```

## Objective

Compare the microbiome of trees with AOD, trees with COD and healthy trees.
The two sites where we have COD samples are Moccas (MOCCA) and Southwick (STHWK)

## Data loading

:::{.callout-important}
### Data filtering and processing
- Only keeping Moccas and Southwick microbiome samples
- removed ASVs with not at least 50 reads (bacteria) or 100 reads (fungi) in one sample
- removed samples with less than 100 reads
- removed ASVs that were not Bacteria or Fungi (including Chloroplasts and Mitochondria)
- updated the 16S taxonomic assignation with GDTP phylogenetic placement
:::

```{r}
#| label: load_data

site_metadata <- read.csv("data/site_metadata.csv")

ps_tot <- purrr::map2(
  c(
    "data/FO_16S_Phyloseq_unfiltered.rds",
    "data/FO_ITS_Phyloseq_unfiltered.rds"
  ),
  c(50, 100),
  \(rds, threshold) {
    ps <- readRDS(rds) |>
      # filtering out spurious ASVs
      subset_taxa(Kingdom %in% c("Fungi", "Bacteria")) |>
      subset_taxa(Order != "Chloroplast") |>
      subset_taxa(Family != "Mitochondria") |>
      filter_taxa(\(x) any(x > threshold), prune = TRUE) |>
      {
        \(x) prune_samples(sample_sums(x) >= 100, x)
      }()
    ps <- prune_samples(
      sample_data(ps)$SampleType == "Sample" &
        sample_data(ps)$SampleOrigin == "Microbiome" &
        sample_data(ps)$Site %in% c("MOCCA", "STHWK") &
        is.na(sample_data(ps)$IsolationMethod),
      ps
    )
    sample_data(ps)$disease_tissue <- paste(
      sample_data(ps)$Disease,
      sample_data(ps)$Tissue,
      sep = "_"
    )
    sample_data(ps)$site_tissue <- paste(
      sample_data(ps)$Site,
      sample_data(ps)$Tissue
    )
    sample_data(ps) <- cbind(
      sample_data(ps),
      site_metadata[match(sample_data(ps)$Site, site_metadata$Site), -1]
    )
    return(ps)
  }
)
names(ps_tot) <- c("bacteria", "fungi")
```

Swapping the 16S taxonomic assignation with the `phyloplace` output from the [`nf-core/ampliseq` pipeline](https://nf-co.re/ampliseq/2.6.1/).

```{r}
#| label: fix_bacteria_taxo
phylo_tax <- read.table("data/asvs.taxonomy.sativa.tsv", sep = "\t")
names(phylo_tax) <- c("ASV_ID", "Lineage", "Support")

phylo_tax <- phylo_tax |>
  mutate(
    Kingdom = stringr::str_extract(Lineage, "d__([[:alnum:]]+)\\;?", group = 1),
    Phylum = stringr::str_extract(Lineage, "p__([[:alnum:]_-]+)\\;?", group = 1),
    Class = stringr::str_extract(Lineage, "c__([[:alnum:]_-]+)\\;?", group = 1),
    Order = stringr::str_extract(Lineage, "o__([[:alnum:]_-]+)\\;?", group = 1),
    Family = stringr::str_extract(Lineage,"f__([[:alnum:]_-]+)\\;?",group = 1),
    Genus = stringr::str_extract(Lineage, "g__([[:alnum:]_-]+)\\;?", group = 1),
    Species = stringr::str_extract(Lineage,"s__([[:alnum:]_-]+ [[:alnum:]_-]+)",group = 1)
  ) |>
  select(ASV_ID, Kingdom:Species) |>
  filter(ASV_ID %in% taxa_names(ps_tot$bacteria)) |>
  as.matrix()
rownames(phylo_tax) <- phylo_tax[,"ASV_ID"]
phylo_tax <- phylo_tax[,-1]

tax_table(ps_tot$bacteria) <- tax_table(phylo_tax[taxa_names(ps_tot$bacteria),])
```

Then I am splitting the datasets per tissue to ease the following analysis

```{r}
#| label: split_tissue_disease
levels_tissue <- sample_data(ps_tot$bacteria)$Tissue |>
  unique() |>
  sort()

levels_tissue_disease <- sample_data(ps_tot$bacteria)$disease_tissue |>
  unique() |>
  sort()

ps_16S_tissue <-
  purrr::set_names(levels_tissue) |>
  purrr::map(function(l) {
    to_keep <- sample_data(ps_tot$bacteria)$Tissue == l
    prune_samples(to_keep, ps_tot$bacteria)
  })

ps_16S_tissue_disease <-
  purrr::set_names(levels_tissue_disease) |>
  purrr::map(function(l) {
    to_keep <- sample_data(ps_tot$bacteria)$disease_tissue == l
    prune_samples(to_keep, ps_tot$bacteria)
  })

ps_ITS_tissue <-
  purrr::set_names(levels_tissue) |>
  purrr::map(function(l) {
    to_keep <- sample_data(ps_tot$fungi)$Tissue == l
    prune_samples(to_keep, ps_tot$fungi)
  })

ps_ITS_tissue_disease <-
  purrr::set_names(levels_tissue_disease) |>
  purrr::map(function(l) {
    to_keep <- sample_data(ps_tot$fungi)$disease_tissue == l
    prune_samples(to_keep, ps_tot$fungi)
  })
```

:::{.panel-tabset}

### 16S dataset summary

```{r}
#| label: full_bacteria_table
#| results: 'asis'
res <- purrr::imap(ps_16S_tissue_disease, function(x, i) {
  phyloseq_summary_table(x, name = i)
}) |>
  bind_rows()

res |>
  knitr::kable(col.names = gsub("\\.", " ", names(res)))
```

### ITS dataset summary

```{r}
#| label: full_fungi_table
#| results: 'asis'
res <- purrr::imap(ps_ITS_tissue_disease, function(x, i) {
  phyloseq_summary_table(x, name = i)
}) |>
  bind_rows()

res |>
  knitr::kable(col.names = gsub("[\\.X]", " ", names(res)))
```

:::

### Exporting ASV tables

```{r}
#| label: export_asv_tables

ps_tot |>
  purrr::iwalk(function(ps, i) {
    asv_tab <- ps |>
      phyloseq::otu_table() |>
      as.data.frame() |>
      tibble::rownames_to_column("ASV_ID")
    asv_sha <- asv_tab$ASV_ID
    asv_tab$ASV_ID <- paste0("ASV_", seq_along(asv_sha))

    tax_tab <- ps |>
      phyloseq::tax_table() |>
      as.data.frame() |>
      tibble::rownames_to_column("ASV_sha")
    
    tax_tab <- tax_tab |>
      tibble::add_column(
        "ASV_ID" = asv_tab$ASV_ID[match(tax_tab$ASV_sha, asv_sha)],
        .before = 1
      )
    sample_tab <- ps |>
      phyloseq::sample_data() |>
      data.frame() |>
      tibble::rownames_to_column("SampleID")

    write.csv(
      asv_tab,
      paste0("data/FO_COD_", i, "_ASV_table.csv"),
      row.names = FALSE
    )

    write.csv(
      tax_tab,
      paste0("data/FO_COD_", i, "_tax_table.csv"),
      row.names = FALSE
    )

    write.csv(
      sample_tab,
      paste0("data/FO_COD_", i, "_sample_table.csv"),
      row.names = FALSE
    )
  })
```

## Ordination plots (for Moccas samples only)

::: {.panel-tabset}

```{r}
#| label: ordination_plots
#| fig-cap: "Ordination plots of the microbiome samples from trees with AOD, COD and healthy trees. The ordination is based on Bray-Curtis dissimilarity and PCoA. The samples are colored by disease status (Healthy, AOD, COD)."
#| results: 'asis'

library(ggplot2)

purrr::iwalk(list(bacteria = ps_16S_tissue, fungi = ps_ITS_tissue), \(ps_list, marker) {
  purrr::iwalk(ps_list, \(ps_tissue, t) {
    ps_tissue <- prune_samples(sample_data(ps_tissue)$Site == "MOCCA", ps_tissue)
    ps.ord <- ordinate(ps_tissue, "PCoA", "bray")
    pord <- plot_ordination(ps_tissue, ps.ord, type = "samples", color = "Disease")
    
    cat('### ', paste(t, marker), '\n\n')
    print(pord + geom_point(size = 5, aes(color = Disease)) + 
      scale_color_manual(values = hue_disease) +
      theme_bw() +
      ggtitle(paste(t, marker)))
    cat('\n\n')
  })
})

```

:::

## Alpha diversity analysis (for Moccas samples only)

::: {.panel-tabset}

```{r}
#| label: alpha_diversity
#| results: 'asis'

purrr::iwalk(list(bacteria = ps_16S_tissue, fungi = ps_ITS_tissue), \(ps_list, marker) {
  purrr::iwalk(ps_list, \(ps_tissue, t) {
    ps_sub <- prune_samples(sample_data(ps_tissue)$Site == "MOCCA", ps_tissue)
    p <- plot_richness(ps_sub, x = "Disease", color = "Disease", measures = c("Chao1", "Shannon")) +
      geom_boxplot(alpha = 0.5, aes(fill = Disease)) +
      scale_color_manual(values = hue_disease) +
      scale_fill_manual(values = hue_disease) +
      theme_bw() +
      ggtitle(paste(marker, t, "samples"))
    cat("### ", paste(t, marker), "\n\n")
    print(p)
    cat("\n\n")
  })
})
```

:::

## Session info

<div style="height:300px;overflow:auto;">
```{r}
sessioninfo::session_info()
```
</div>


